A non-additive generalization of probability and \\ state count for all physical theories

### 2025-12-23 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: more trimming

Since physical laws deal with experimentally reproducible relationships, they are inherently relationships between statistical ensembles. Therefore, any physical theory must define the space of all ensembles allowed within the theory. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (e.g. a fair coin) or over phase space (e.g. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (e.g. a positive semi-definite self-adjoint operator of trace one).

In this work, we will take an ensemble space $\Ens$ to be a bounded convex subset of a locally convex metrizable second countable topological vector space endowed with an entropy function. The topology captures experimental verifiability as open sets represent experimentally verifiable statements\cite{aop-book,kelly1996}.\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Convexity, and linearity more in general, comes from allowing statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i=1}^{\infty} p_i \ens_i$ converges in the space or not.} Lastly, the entropy characterizes the variability of the elements within an ensemble. In this context, a statistical quantity, representing the expectation of a real valued quantity, is an affine continuous function $F : \Ens \to \mathbb{R}$. That is, $F(p \ens[a] + \bar{p} \ens[b]) = p F(\ens[a]) + \bar{p} F(\ens[b])$.

Each physical theory satisfies these axioms, even though it may be represented very differently mathematically. Classical mechanics, for example, is typically described in terms of the geometry of phase space. Quantum mechanics is most often defined in terms of complex inner product spaces, but can be reframed in terms of quasiprobability distributions. Therefore, while the above requirements are needed for a physical theory, it does not necessarily mean that it is convenient to describe problems or make calculations in terms of those definitions. In fact, those requirements interact with each other leading to a rich mathematical structure that connects to many of the structures we use in mathematical physics, and our goal is mainly to see how the different representations are connected. In particular, here we want to examine the different ways that measure theory can be used to represent ensembles in any given physical theory.

\textbf{Ensemble spaces as convex sets.} For the purpose of measure theoretic representations, it is best to understand an ensemble space $\Ens$ as a convex subset of a metrizable second countable locally convex topological vector space $V$.\footnote{To see the connection between the requirements and this mathematical characterization see the appendix.} See Fig.~\ref{convex_examples} for examples. The entropy is a function $S : \Ens \to \mathbb{R}$ that satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + \bar{p} &S(\ens[b]) \leq S(p \ens[a] + \bar{p} \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + \bar{p} S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{The Shannon entropy is recovered as the only upper bound possible.}  Intuitively, the variability (i.e.~the entropy) cannot decrease during mixing, and will have an upper bound when mixing two ensembles whose instances are always distinguishable.

Using the convex structure, we say that two ensembles are separate, noted $\ens[a] \separate \ens[b]$, if they cannot be expressed as a mixture of a common ensemble. That is, there is no $\ens[c]$ such that $\ens[a] = \alpha \ens[c] + \bar{\alpha} \ens[d]$ and $\ens[b] = \beta \ens[c] + \bar{\beta} \ens$ from some $\alpha, \beta \in (0,1]$ and $\ens[d],\ens \in \Ens$. No ensemble is separate from itself. An internal point is an ensemble that is not separate from any ensemble (i.e.~it is the mixture of any ensemble). An extreme point is an ensemble that is separate from all others (i.e.~it is only a mixture of itself).

@@

		% meridians and equator
		\begin{scope}[canvas is yz plane at x=0]
			\draw (\ixy, \ixz) arc[start angle={atan2(\ixz, \ixy)}, delta angle=180, radius=1 cm];
			\draw[dashed] (-\ixy, -\ixz) arc[start angle={atan2(-\ixz, -\ixy)}, delta angle=180, radius=1 cm];
		\end{scope}
		\begin{scope}[canvas is zx plane at y=0]
			\draw[dashed] (\iyz, \iyx) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=1 cm];
			\draw (-\iyz, -\iyx) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=1 cm];
		\end{scope}
		\begin{scope}[canvas is xy plane at z=0]
			\draw[dashed] (\izx, \izy) arc[start angle={atan2(\izy, \izx)}, delta angle=180, radius=1 cm];
			\draw (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, delta angle=180, radius=1 cm];
		\end{scope}
	\end{tikzpicture}
	\caption{Examples of a classical discrete ensemble space with three pure states (triangle) and of a two dimensional quantum ensemble space (ball).}\label{convex_examples}
\end{figure}

The set of all extreme points $X$, as we will see, plays a major role. However, on physical grounds, these may not be proper ensembles. In classical mechanics, these would correspond to Dirac measures: probability distributions all concentrated at a single point of phase space. Since we cannot reliably prepare a particle with infinite precision, and its entropy would correspond to minus infinity, these are not ensembles and they are not part of $\Ens$. For the purpose of Choquet theory, we will assume that there is a way to find the closure $E \supset \Ens$ that includes all extreme points $X$. Moreover, we will also assume $E$ to be compact, even though it will not be in general. How to properly make these identifications, and how to go beyond the compact case, is outside the scope of this particular article.\footnote{An early attempt used the notion of orthogonality to create the lattice of orthogonal subspaces and use something along the lines of Stone's representation theorem for Boolean algebra. This does not work. An approach, currently under investigation, is to study limits over statistical quantities. We are still trying to understand what mathematical structures have enough information to allow the reconstruction.} Note that the classical discrete and the quantum case fall in the compact case. Moreover, the set of all probability distributions over a compact space $X$ also falls in the compact case. Therefore, apart from the technicality above, the compact case is still a very interesting case. To distinguish the context, in this article we will use $\Ens$ to indicate the general case and $E$ to indicate this ``compact closure.''

@@

To summarize, a repeatedly verifiable physical theory requires ensembles, which must be defined experimentally, allow statistical mixing and provide an entropy. Additionally, we require the existence of enough statistical quantities to identify each ensemble. Under these assumptions, an ensemble space $\Ens$ is a convex set of a metrizable second countable locally convex topological vector space $V$. In some cases, for simplicity, we will require $\Ens$ to admit a ``compact closure'' $E$, so that it includes the set $X$ of its extreme points. We now turn to the question of how we can represent both the entropy and each ensemble in terms of some measure theoretic structure.

@@

\textbf{Expectation and integration.} A random variable is a function $F : X \to \mathbb{R}$ that assigns a value to each state.\footnote{The function $F$ must be measurable, in the sense that the inverse of a Borel set should be a Borel set. We find that the term is confusing for a physics audience, so it is avoided in the main text.} The number of the side of the die, position, and momentum are typical examples. The expectation for a random variable is given by the integral
\begin{equation}
	E[F] = \int_X F dp.
\end{equation}
The expectation is linear with respect to the variable (i.e.~$E[aF+bG]= aE[F]+bE[G]$). In our case, what is more interesting is that it is an affine of the probability measure: if $p = \lambda p_1 + \bar{\lambda} p_2$ then $\int_X F dp = \lambda \int_X F dp_1 + \bar{\lambda} \int_X F dp_2$. That is, as we required for statistical variables, the expectation for a mixture is the mixture of the expectations.

@@

>>>>

@@

		% meridians and equator
		\begin{scope}[canvas is yz plane at x=0]
			\draw (\ixy, \ixz) arc[start angle={atan2(\ixz, \ixy)}, delta angle=180, radius=1 cm];
			\draw[dashed] (-\ixy, -\ixz) arc[start angle={atan2(-\ixz, -\ixy)}, delta angle=180, radius=1 cm];
		\end{scope}
		\begin{scope}[canvas is zx plane at y=0]
			\draw[dashed] (\iyz, \iyx) arc[start angle={atan2(\iyx, \iyz)}, delta angle=180, radius=1 cm];
			\draw (-\iyz, -\iyx) arc[start angle={atan2(-\iyx, -\iyz)}, delta angle=180, radius=1 cm];
		\end{scope}
		\begin{scope}[canvas is xy plane at z=0]
			\draw[dashed] (\izx, \izy) arc[start angle={atan2(\izy, \izx)}, delta angle=180, radius=1 cm];
			\draw (-\izx, -\izy) arc[start angle={atan2(-\izy, -\izx)}, delta angle=180, radius=1 cm];
		\end{scope}
	\end{tikzpicture}
	\caption{Examples of a classical discrete ensemble space with three pure states (triangle) and of a two dimensional quantum ensemble space (ball). Infinite dimensional spaces, like continuous classical ensemble spaces and quantum ensemble space with continuous quantities cannot be visually represented. }\label{convex_examples}
\end{figure}

Since physical laws deal with experimentally reproducible relationships, they are inherently relationships between statistical ensembles. Therefore, any physical theory must define the space of all ensembles allowed within the theory. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (e.g. a fair coin) or over phase space (e.g. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (e.g. a positive semi-definite self-adjoint operator of trace one). See Fig.~\ref{convex_examples} for examples.

In this work, we will take an ensemble space $\Ens$ to be a bounded convex subset of a locally convex metrizable second countable topological vector space endowed with an entropy function. The second countable topology captures experimental verifiability as open sets represent experimentally verifiable statements\cite{aop-book,kelly1996}.\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Convexity, and linearity more in general, comes from allowing statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i=1}^{\infty} p_i \ens_i$ converges in the space or not.} Lastly, the entropy characterizes the variability of the elements within an ensemble. The entropy is a function $S : \Ens \to \mathbb{R}$ that satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + \bar{p} &S(\ens[b]) \leq S(p \ens[a] + \bar{p} \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + \bar{p} S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{The Shannon entropy is recovered as the only upper bound possible.}  Intuitively, the variability (i.e.~the entropy) cannot decrease during mixing, and will have an upper bound when mixing two ensembles whose instances are always distinguishable.

In this context, a statistical quantity, representing the expectation of a real valued quantity, is an affine continuous function $F : \Ens \to \mathbb{R}$. That is, $F(p \ens[a] + \bar{p} \ens[b]) = p F(\ens[a]) + \bar{p} F(\ens[b])$. We also assume that there are enough statistical quantities to identify each ensemble. That is, there is a family $F_i$ such that $\ens[a] = \ens[b]$ if and only if $F_i(\ens[a]) = F_i(\ens[b])$ for all $i \in I$. This connects to the typical constructions for function spaces and their dual.

Since any physical theory must provide experimental verifiability, statistical mixing and an entropy function, this gives us a general and solid conceptual foundation. While we are still in the process of understanding exactly how these requirements give us local convexity and metrizability, for the purpose of this work, we will simply assume them.

\textbf{Extreme points and compactness.} Using the convex structure, we say that two ensembles are separate, noted $\ens[a] \separate \ens[b]$, if they cannot be expressed as a mixture of a common ensemble. That is, there is no $\ens[c]$ such that $\ens[a] = \alpha \ens[c] + \bar{\alpha} \ens[d]$ and $\ens[b] = \beta \ens[c] + \bar{\beta} \ens$ from some $\alpha, \beta \in (0,1]$ and $\ens[d],\ens \in \Ens$. No ensemble is separate from itself. An internal point is an ensemble that is not separate from any ensemble (i.e.~it is the mixture of any ensemble). An extreme point is an ensemble that is separate from all others (i.e.~it is only a mixture of itself).

The set of all extreme points $X$, as we will see, plays a major role in Choquet theory. However, on physical grounds, these may not be proper ensembles. In classical mechanics, these would correspond to Dirac measures: probability distributions all concentrated at a single point of phase space. Since we cannot reliably prepare a particle with infinite precision, and its entropy would correspond to minus infinity, these are not ensembles and they are not part of $\Ens$. For the purpose of Choquet theory, we will assume that there is a way to find the closure $E \supset \Ens$ that includes all extreme points $X$. Moreover, we will also assume $E$ to be compact, even though it will not be in general. How to properly make these identifications, and how to go beyond the compact case, is outside the scope of this particular article.\footnote{An early attempt used the notion of orthogonality to create the lattice of orthogonal subspaces and use something along the lines of Stone's representation theorem for Boolean algebra. This does not work. An approach, currently under investigation, is to study limits over statistical quantities. We are still trying to understand what mathematical structures have enough information to allow the reconstruction.} Note that the classical discrete and the quantum case fall in the compact case. Moreover, the set of all probability distributions over a compact space $X$ also falls in the compact case. Therefore, apart from the technicality above, the compact case is still a very interesting case. To distinguish the context, in this article we will use $\Ens$ to indicate the general case and $E$ to indicate this ``compact closure.''

@@

To summarize, a repeatedly verifiable physical theory requires ensembles, which must be defined experimentally, allow statistical mixing and provide an entropy. Additionally, we require the existence of enough statistical quantities to identify each ensemble. We therefore take an ensemble space $\Ens$ to be a bounded convex subset of a metrizable second countable locally convex topological vector space $V$. In some cases, for simplicity, we will require $\Ens$ to admit a ``compact closure'' $E$, so that it includes the set $X$ of its extreme points. We now turn to the question of how we can represent both the entropy and each ensemble in terms of some measure theoretic structure.

@@

\textbf{Expectation and integration.} A random variable is a function $F : X \to \mathbb{R}$ that assigns a value to each state.\footnote{The function $F$ must be measurable, in the sense that the inverse of a Borel set should be a Borel set. We find that the term is confusing for a physics audience, so it is avoided in the main text.} The number of the side of the die, position, and momentum are typical examples. The expectation for a random variable is given by the integral
\begin{equation}
	E[F] = \int_X F dp.
\end{equation}
The expectation is linear with respect to the variable (i.e.~$E[aF+bG]= aE[F]+bE[G]$). In our case, what is more interesting is that it is an affine function of the probability measure: if $p = \lambda p_1 + \bar{\lambda} p_2$ then $\int_X F dp = \lambda \int_X F dp_1 + \bar{\lambda} \int_X F dp_2$. That is, as we required for statistical variables, the expectation for a mixture is the mixture of the expectations.

@@

TODO add picture
